---
title: "CCHMC R Users Group"
subtitle: "Crash Course in R"
author: "Chris Wolfe, MS"
date: "`r Sys.Date()`"

output:
  xaringan::moon_reader:
    self_contained: true
    seal: false
    nature:
      scroll: false
      click: false
      touch: false
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
options(smart = TRUE)
knitr::opts_chunk$set(comment = '')
```

```{r echo = FALSE, message = FALSE}
library(tidyverse)
```


class: center, middle

# Crash Course in R

### CCHMC R Users Group
### Cincinnati, OH

#### Chris Wolfe, MS
#### `r Sys.Date()`

---
class: center, middle

## **DISCLAIMER**: this is not an attempt to convert you if you are a SAS user (*although we are willing to help should you choose to do so*).

---

# Introduction


The purpose of this group is to facilitate interaction among anyone using R within the Academic Health Center with the goal to increase knowledge and collaboration among its members.


### Help us get to know you

Who are you? What is your experience with R? What do you hope to learn?

---

# Where can I learn R?

#### There are numerous other books and online programs, but here are some good places to start:

- Coursera (https://www.coursera.org/)
- R for Data Science (https://r4ds.had.co.nz/)
- Hands on R Programming (https://rstudio-education.github.io/hopr/)
- R for Dummies 

Whatever type of learner you are, you are bound to find a suitable option.

---
# What is R?

R is a language and environment for statistical computing and graphics. 

#### History

- R was developed in the early 1990s at Bell Laboratories, and is based on the statistical programming language "S".
- In 1995, R was made available as a free software.
- The first version of R (Version 1.0.0) was realeased.
- The current version of R is v3.61

#### Functional Programming

- R is a functional programming language--it provides many tools for the creation and manipulation of functions.
- Data manipulation and analysis with R typically start with small, easy-to-understand building blocks, wcich are combined to form more complex structures to achieve a desired goal

.footnote[https://www.r-project.org/about.html]
---

# Why use R?

#### Advantages:

- **R is open-source and free to use**
    + Improvements and fixes by programmers keeps R stable, reliable, has vastly expanded its capabilities
- **Runs on multiple OS platforms (Mac, Windows, Linux)**
    + Not true for SAS...point for R!
- **Supports extensions (packages)**
    + R is powerful by itself, but its capabilities can be extended by additional software packages
    + Often times, statistical methods are published in parallel with an associated R package
- **Large, active user community**
    + Blogs, Q&A sites, regional and international conferences
- **Connection with other languages**
    + Data from other software can be imported and exported (i.e SAS)
    + SQL database connection
    
---

# Code Editing

There are multiple code editors or graphical user interfaces (GUI) to write R code and perform analyses.

#### R Gui
- This is the standard code editor that accompanies the R install
- Very basic--includes the console window where code is written

#### R Studio
- Highlights code to differentiate keywords and variables (easier to read)
- Automatic bracket and parenthesis matching
- Code autocompletion
- Shows other windows beyond console (source, environment, plots/help/files)
- Capability to create multiple files of different formats (.R, .Rmd, .txt, .pdf, .html, .pptx)
- Easily manage multiple files and directories by creating "projects"

---

## Data *types* and *structures* in R

**R has 6 basic data types:**

- character: `"a"`, `"RUG"`
- numeric: `3`, `100.52`
- integer: `5L` (the `L` tells R to store this as an integer)
    + factors are a type of integer
- logical: `TRUE`, `FALSE`
- complex: `1+8i` (complex numbers with real and imaginary parts)

Elements of the different data types can be combined to form data *structures.*

**R has 5 basic data structures:**

- atomic vector
- matrix
- list
- data frame
- array

R data structures can be categorized by their dimensionality and whether their contents are homogeneous.

---

# Atomic Vectors

The moste basic data structure in R is the vector. There are two types of vectors: *lists* and *atomic vectors*.  Typically, the term "vector" usually refers to an atomic vector.  Vectors are of a single dimension.

**All elements of an atomic vector must be of the same type**. 

There are a few base functions that R provides that examines the features of vectors (and other data structures):

```{r}
x <- c("a", "b", "ab", "d") ## THIS CREATES A NUMERIC VECTOR OF THE NUMBERS 1- 5
                            # c() IS A FUNCTION THAT COMBINES -
length(x)                   # ARGUEMENTS TO FORM A VECTOR
str(x)

```
---

# Lists

Lists are different from atomic vectors because elements can be of **any** type, including lists. You construct lists by using list(). Lists are of one dimension.

```{r}
x <- list(1:3, "a", c(2.3, 5.9)) 
str(x)

```

Lists can be recursive vectors, meaning a list can contain other lists. 

```{r}
x <- list(list(num = 1:3), c(TRUE, FALSE, TRUE)) 
str(x)
```

---

# Matrices and arrays

Adding a `dim()` (i.e. dimensionality) attribute to an atomic vector allows it to behave like a multi-dimensional array. A special case of the array is the matrix, which has two dimensions. Matrices can also be created by calling the `matrix()` function.

```{r}
a <- matrix(1:8, ncol = 4, nrow = 2)
a
c <- 1:10
dim(c) <- c(2, 5)
c

```

---

# Data Frames

*A data frame is the most common way of storing data in R.* A data frame is a list of equal-length vectors and has a 2-dimensional structure, so it shares properties of both the matrix and the list. In other words, it is a 'rectangular' list.

The `data.frame()` function can be used to create a data frame by hand, although this is typically not useful.

```{r}
x <- data.frame(id = letters[1:3], x = 1:3, y = 11:13)
```

#### `$`

The `$` is used to access specific variables of a data frame--R will return a vector of all the values contained in the variable. 

```{r}
x <- head(mtcars$cyl)
x
```

---

# Data Frames

Like `str()` and `length()` there are other useful functions that help you to examine the features of data frames.

- `ncol()` outputs the number of columns
- `nrow()` outputs the number of columns
- `glimpse()` allows you to previous your data like `str()` but the output is a bit cleaner

```{r}
x <- data.frame(id = letters[1:3], x = 1:3, y = 11:13)
glimpse(x)
```

---

class: center, middle

# Coding Basics

---

## Calculation

R can do calculations on the fly in the console:

```{r}
1 / 200 * 30
(59 + 73 + 2) / 3
2 + 2^5 - (100*2)
```

---

## Assigning Values

New R objects can be created using the `<-` or the *assignment operator*.

All R statements where you create objects, or assignment statements, have the same form:

```{r}
"object_name <- value"
```

**`=` can also be used to assign values to R objects, but is not recommended.**

Using `=` can cause confusion down the line. 

RStudio’s keyboard shortcut for the assingment operator is: Alt + - (the minus sign).  If you are using a Mac, the shortcut is: option + - (the minus sign).

---

## Object Names

Object names must start with a letter, and can only contain letters, numbers, `_` and `.`. You want your object names to be descriptive, so you’ll need a convention for multiple words. 

```{r eval = FALSE, echo = TRUE}
"valid_name" 
"0_not_valid"
"this.works" 
"this_also.works_BUT.isconfusingandlong"
```

You can inspect an object by typing its name (if you hadn't caught on from prevous slides).

---

## Calling Functions

R has a large collection of built-in functions that are called like this:

```{r, eval = FALSE, echo = TRUE}
"function_name(arg1 = val1, arg2 = val2, ...)"
```

```{r}
x <- 1:100

mean(x)
sd(x)
summary(x)
```

---
## Vectorization

*R is a vector-based language.*

Vectorization allows you to perform sometimes complex operations on a set of values with a single command. Calculations involving vectors occur over the length of the entire vector

```{r}
x <- 1:10
x + 2
```

Vectors can also be added to another vector.

```{r}
x <- c(5, 1, 4, 5, 9, 0)
x + 6:11
```

---
## Missing Data

Missing data is represented by `NA` in R.

```{r}
x <- c("a", NA, "c", "d", NA)
x
```

The function is.na() indicates the elements of the vectors that represent missing data and returns a *logical* vector.

```{r}
x <- c("a", NA, "c", "d", NA)
is.na(x)
```

---

## Helpful Functions

It is impossible to name all of the helpful functions in R, but here are a few:

- `getwd()` prints the name of your current working director
- `setwd()` changes your working directory
- `list.files()` returns a character vector of all files in your working directory
- `unique()` returns a vector of unique values
- `duplicated()` returns a logical vector equal to the length of the input
- `table()` returns an obect of class `table` that shows the number of observations within each level of a factor
- `paste()` concatenate vectors after converting to character
- `range()` returns the minimum and maximum 
- `summary()` returns summary table with minimum and maximum values, mean, and quartiles of a numeric vector
- `Sys.date()` returns current date
- `Sys.time()` returns current date-time

---

## User Defined Functions

Functions allow you to automate common tasks in a more powerful and general way than copy-and-pasting. Functions have three major advantages over using copy-and-paste:

- You can give a function a name that makes your code easier to understand
- As requirements change, you only need to update code in one place, instead of many
- You eliminate the chance of making incidental mistakes when you copy and paste (i.e. updating a variable name in one place, but not in another).

**General rule of thumb:** You should consider writing a function whenever you’ve copied and pasted a block of code more than twice.

The basic anatomy of a structure looks like this:

```{r}
myfunction <- function(arg1, arg2, ... ){
    statements
    return(object)
}
```

---

# Installing Packages

What is an "R package?"
- a package is a collection of code, help files, datasets and other files that can be incorporated into R and expand its capabilities
- packages exist for new statistical methods, emojis, data manipulation

There is real power in the fact that any R user can write and contribute code and share their package in an organized manner.

**If you need to do something specific, chances R that a package exists** (or something close to it that you can expand on).

Packages can be installed using the `install.packages()` function.

```{r, eval = FALSE}
install.packages("tidyverse")
```

Packages can be loaded using the `library()` function.

```{r, eval = FALSE}
library(tidyverse)
```

